---
title: "Week 03 Assignment"
author: "Harrison Fuller"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    citation_package: natbib
    latex_engine: pdflatex
    template: /Users/harrison.fuller/OneDrive - Washington State University (email.wsu.edu)/Classes/STATS
      419/WSU_STATS419_FALL2020/latex-templates/homework.tex
course: STATS 419 Survey of Multivariate Anlaysis
course-short: STATS 419
instructor: Monte J. Shaffer
email: harrison.fuller@wsu.edu
fig_caption: yes
number_sections: yes
myVar: 'This is a variable'
---


``````{r, mychunk-path, message=FALSE}
library(devtools);
github.path = "https://raw.githubusercontent.com/fullerharrison/WSU_STATS419_FALL2020/";
source_url(paste0(github.path, "master/functions/libraries.R"))
```


# Matrix

Create the "rotate matrix" functions described in lectures from the sample matrix.  Apply to the example "myMatrix".


```{r, mychunk-matrix-include, message=FALSE}
source_url(paste0(github.path, "master/functions/functions_matrix.R"));

myMatrix = matrix(c(
  1, 0, 2,
  0, 3, 0,
  4, 0, 5), nrow = 3, byrow = T)

```

```{r, mychunk-matrix-demo}

transposeMatrix(myMatrix);
rotateMatrix90(myMatrix); # clockwise
rotateMatrix180(myMatrix);
rotateMatrix270(myMatrix)
```